Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-16 Manuscript/manuscript.tex
Removed a few red lines -- mostly checks

According to ECC mean-field correction\todo{HECTOR: I added this back because it is an important detail.}, the electronic polarizability can be modeled in non-polarizable MD by effectively embedding the ions in a homogeneous dielectric continuum with a dielectric constant $\epsilon_{el}$, which is the electronic part of the dielectric constant of the medium~\cite{leontyev11}. Following Coulomb's law,  ECC can be directly incorporated by scaling the charges with a constant scaling factor of $f_q = \epsilon _{el} ^{-1/2}$, giving
\begin{equation}
Q^{ECC} = f_q \cdot Q
\end{equation}
for the ECC corrected charges in simulation. 
Given that the  high frequency dielectric constant of water is $\epsilon _{el} = 1.78$ (\textit{i.e.,} the square of the refraction index), the scaling factor for ions in water is roughly $f_q \approx 0.75$. This scaling factor successfully improves the accuracy of simulations of solvated ions, when quantitatively compared with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014, martinek17}.
It is important to note that the 
high frequency dielectric constant 
is around 2 for almost any biologically relevant environment \cite{leontyev11}.
The dielectric discontinuity in a lipid bilayer arises
from the orientational polarization of the molecules. 
Thus, the same correction for the electronic polarizability can be 
applied throughout the lipid bilayer interface.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules
and approximately 50~water molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.
The SPC/E~\cite{Berendsen1987} water model was used in all ECC-POPC model simulations reported
in the main text. This model was also used in the previous parametrization
of ECC-ions \cite{martinek17, kohagen16, Pluharova2014} because its lowered dielectric constant is consistent with
the ECC concept~\cite{leontyev11, leontyev14}.
The robustness of our approach with other water models (OPC \cite{Izadi14},
OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005})
is tested in the SI.
Sodium, calcium, and chloride ions were modeled as ECC-ions
with parameters from Refs. \citenum{martinek17, Pluharova2014}. Scaled charges and Lennard-Jones radii for atoms
forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}.
For comparison, simulations with the Lipid14 model~\cite{dickson14} and
ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} or ECC-ions \cite{martinek17, kohagen16, Pluharova2014}
were also performed. The TIP3p water model~\cite{jorgensen83} was used in all simulations with the original Lipid14 model.
\todo{Joe, please confirm that this is correct.}
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at {\color{red}??K}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at {\color{red}??K}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
    {\color{red}Joe, please add the temperatures}
  } 
\end{figure}

@@

\begin{table}[tb!]
  \caption{Bulk concentrations of Ca$^{2+}$ (C$_b$), relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$),
    and the percetages of Ca$^{2+}$ bound to phosphate or carbonyl oxyges ($r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ and $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$)
    in different POPC bilayer models. All systems have the same molar concentration of Ca$^{2+}$ respect to water ($C_{ion}'$=350mM).
    %{\color{red}HECTOR:PLEASE CHECK}
    {\color{red}Joe, please confirm that $r^\mathrm{Ca^{2+}} _\mathrm{PO_4}$ is really 100\% for all the Lipid14 simulations
      (i.e. that this is not a typo).}
  \label{tab:binding}}
  \begin{tabular}{l|c c | c | c c}
    model                  & $C_{ion}'$ & $C_{ion}\,/\,\mathrm{mM}$ & $\Gamma_{Ca}^{\rm water} \,/\, \mathrm{nm}^{-2}$  & $r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-POPC             &  350  &  $280\pm 10 $  &  $0.06 \pm 0.01 $                           &  99\%  &    25\%    \\
    Lipid14/\AA{}qvist     &  350  &  $210\pm 10 $  &  $0.13 \pm 0.01 $                          & 100\%  &    37\%     \\
    Lipid14/Dang           &  350  &  $160\pm 10 $  &  $0.23 \pm 0.03 $                            & 100\%  &    14\%    \\
    Lipid14/ECC-ions       &  350  &  $120\pm 100$  &  $0.35 \pm 0.11 $                         & 100\%  &    23\%    \\
%Average no. of bound cations:
%L14+Aqvist: all:22.7, phos:22.7, carb:8.5
%L14+Dang:   all:29.2, phos:29.2, carb:4.0
%L14+ECC-ion:all:37.7, phos:37.7, carb:8.5
  \end{tabular}
\end{table}

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields.
    %Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    {\color{red} Joe, please clarify the sentence below}
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    densities of~\ce{Cl^-} and \ce{Na^{+}} are divided by 2, and
    densities of phosphate group resp. water are divided by 5 resp. 200. 
    All the simulations shown here have the same molar concentration of ions in water ($C_{ion}'$=350~mM).
    }
\end{figure}

@@

\todo{HECTOR: I do fully understand the content of this paragraph. I will discuss this with Joe tomorrow. Please be carefull with my changes.} The histograms of residence times of \ce{Ca^{2+}} ($C_{ion}'$ = 450~mM) in a POPC bilayer for simulations using either ECC-POPC or CHARMM36 (Simulation from Refs.~\citenum{javanainen17,zenodo.259376}) are shown in Fig.~S4 in SI. In the CHARMM36 simulation, a significant number of the calciums are bound for the whole length of the trajectory ({\color{red} 800~ns}). In contrast, at least an order of magnitude faster bound/unbound calcium exchange is observed within the ECC-POPC model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are shorter than $60\,\mathrm{ns}$. The longest observed residence time is around 150~ns, which is well below the total length of the simulation used for analysis, i.e., 200~ns. Note that these results are in line with the experimental estimate that the residence time of \ce{Ca^{2+}} at each PC headgroup is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}. The exchange of \ce{Na+} ions at the POPC membrane is another order of magnitude faster, yielding 90\% of the residence times smaller than~1~ns, with the longest residence time being~6~ns.

>>>>

According to ECC, the electronic polarizability can be modeled in non-polarizable MD in a mean-field way by 
embedding the ions in a homogeneous dielectric continuum with a dielectric constant $\epsilon_{el}$, which is the electronic part of the dielectric constant of the medium~\cite{leontyev11}. Following Coulomb's law,  ECC can be directly incorporated by scaling the charges with a constant scaling factor of $f_q = \epsilon _{el} ^{-1/2}$, giving
\begin{equation}
Q^{ECC} = f_q \cdot Q
\end{equation}
for the ECC corrected charges in simulation. 
Given that the  high frequency dielectric constant of water is $\epsilon _{el} = 1.78$ (\textit{i.e.,} the square of the refraction index), the scaling factor for ions in water is roughly $f_q \approx 0.75$. This scaling factor successfully improves the accuracy of simulations of solvated ions, when quantitatively compared with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014, martinek17}.
It is important to note that the 
high frequency dielectric constant 
is around 2 for almost any biologically relevant environment \cite{leontyev11}.
The dielectric discontinuity in a lipid bilayer arises
from the orientational polarization of the molecules. 
Thus, the same correction for the electronic polarizability can be 
applied throughout the lipid bilayer interface.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules
and approximately 50~water molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.
The SPC/E~\cite{Berendsen1987} water model was used in all ECC-POPC model simulations reported
in the main text. This model was also used in the previous parametrization
of ECC-ions \cite{martinek17, kohagen16, Pluharova2014} because its lowered dielectric constant is consistent with
the ECC concept~\cite{leontyev11, leontyev14}.
The robustness of our approach with other water models (OPC \cite{Izadi14},
OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005})
is tested in the SI.
Sodium, calcium, and chloride ions were modeled as ECC-ions
with parameters from Refs. \citenum{martinek17, Pluharova2014}. Scaled charges and Lennard-Jones radii for atoms
forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}.
For comparison, simulations with the Lipid14 model~\cite{dickson14} and
ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006} or ECC-ions \cite{martinek17, kohagen16, Pluharova2014}
were also performed. The TIP3p water model~\cite{jorgensen83} was used in all simulations with the original Lipid14 model.
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at 303~K.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at 300~K.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

@@

\begin{table}[tb!]
  \caption{Bulk concentrations of Ca$^{2+}$ (C$_b$), relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$),
    and the percetages of Ca$^{2+}$ bound to phosphate or carbonyl oxyges ($r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ and $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$)
    in different POPC bilayer models. All systems have the same molar concentration of Ca$^{2+}$ respect to water ($C_{ion}'$=350mM).
    %{\color{red}HECTOR:PLEASE CHECK}
    %{\color{red}Joe, please confirm that $r^\mathrm{Ca^{2+}} _\mathrm{PO_4}$ is really 100\% for all the Lipid14 simulations
    %  (i.e. that this is not a typo).}
  \label{tab:binding}}
  \begin{tabular}{l|c c | c | c c}
    model                  & $C_{ion}'$ & $C_{ion}\,/\,\mathrm{mM}$ & $\Gamma_{Ca}^{\rm water} \,/\, \mathrm{nm}^{-2}$  & $r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-POPC             &  350  &  $280\pm 10 $  &  $0.06 \pm 0.01 $                           &  99\%  &    25\%    \\
    Lipid14/\AA{}qvist     &  350  &  $210\pm 10 $  &  $0.13 \pm 0.01 $                          & 100\%  &    37\%     \\
    Lipid14/Dang           &  350  &  $160\pm 10 $  &  $0.23 \pm 0.03 $                            & 100\%  &    14\%    \\
    Lipid14/ECC-ions       &  350  &  $120\pm 100$  &  $0.35 \pm 0.11 $                         & 100\%  &    23\%    \\
%Average no. of bound cations:
%L14+Aqvist: all:22.7, phos:22.7, carb:8.5
%L14+Dang:   all:29.2, phos:29.2, carb:4.0
%L14+ECC-ion:all:37.7, phos:37.7, carb:8.5
  \end{tabular}
\end{table}

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields.
    %Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    %In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    The density profiles of~\ce{Cl^-} and \ce{Na^{+}} ions are divided by 2, and
    the density profiles of phosphate groups resp. water are divided by 5 resp. 200. 
    All simulations with \ce{CaCl2} shown here have the same molar concentration of ions in water ($C_{ion}'$=350~mM).
    The simulation with \ce{NaCl} has $C_{ion}'$=1000~mM.
    }
\end{figure}

@@

\todo{HECTOR: I do fully understand the content of this paragraph. I will discuss this with Joe tomorrow. Please be carefull with my changes.} 
The histograms of residence times of \ce{Ca^{2+}} ($C_{ion}'$ = 450~mM) in a POPC bilayer for simulations using either ECC-POPC or CHARMM36 (Simulation from Refs.~\citenum{javanainen17,zenodo.259376}) are shown in Fig.~S4 in SI. In the CHARMM36 simulation, a significant number of the calciums are bound for the whole length of the trajectory (800~ns). In contrast, at least an order of magnitude faster bound/unbound calcium exchange is observed within the ECC-POPC model, where 90\% of the \ce{Ca^{2+}} residence times to a POPC membrane are shorter than $60\,\mathrm{ns}$. The longest observed residence time is around 150~ns, which is well below the total length of the simulation used for analysis, i.e., 200~ns. Note that these results are in line with the experimental estimate that the residence time of \ce{Ca^{2+}} at each PC headgroup is shorter than $10\,\mu\mathrm{s}$~\cite{altenbach84}. The exchange of \ce{Na+} ions at the POPC membrane is another order of magnitude faster, yielding 90\% of the residence times smaller than~1~ns, with the longest residence time being~6~ns.
